---
title: 'Day 3, exercise 2: Extending to cause specific results'
output: pdf_document
date: '2022-05-07'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In exercise 4, we decomposed life expectancy and lifespan disparity by age. Now, we will extend the decomposition by also considering cause of death. We will use the same data, packages and functions as before.

Let's load them.
```{r, message=FALSE}
load('AburtoBeltranSanchez.RData')

library(DemoDecomp)
library(tidyverse)

source("Functions_D3.R")
```

Following the method, we require two matrices of rates: one for 2005 and one for 2015, both by age (rows) and cause of death (column)
```{r}
#we need age-cause specific mortality rates in each period
COD1 <- data %>% 
  filter(year==2019) %>% 
  select(Cause_1:Cause_9) %>% 
  as.matrix()

COD2 <- data %>% 
  filter(year==2015, age>15) %>% 
  select(Cause_1:Cause_9) %>% 
  as.matrix()
```


We cannot directly use the same functions to calucalte life expectancy and lifespan disparity. Rather, we need functions that can estimate life expectancy and lifespan dispairty from age- and cause-specific death rates. They are also included in the functions script. Let's have a look.

```{r}
e0.frommxc

edagger.frommxc
```

Now we can perform the decomposition and check the results

```{r}
results <- horiuchi(func = e0.frommxc, pars1 = c(COD1), pars2 = c(COD2), N = 100)

#Go back to a matrix
dim(results) <- dim(COD1)
results

#original
(original <- e0.frommxc(COD2) - e0.frommxc(COD1))

#with decomp
(with.decomp <- sum(results))

#error
with.decomp - original
```
Do some data handling to have a nice dataset.

```{r}
results <- data.frame(results)
colnames(results) <- cause_names
results$age <- 16:109
results <- gather(data = results,key = cause,value = contribution,-age)
```

Finally, graph and interpret the results.

```{r}
#now graph results
ggplot(data=results, aes(x=as.factor(age), y=contribution, fill=cause)) +
  ggtitle(bquote(~'Change in '~ e[15] ~'2005-2015' )) +
  geom_bar(stat = "identity", position = "stack")

```


Now for lifespan disparity

```{r}
#Now we can perfom the decomposition
results_edag <- horiuchi(func = edagger.frommxc, pars1 = c(COD1), pars2 = c(COD2), N = 100)
```

Go back to a matrix

```{r}
#Go back to a matrix
dim(results_edag) <- dim(COD1)

#original
(original_edag <- edagger.frommxc(COD2) - edagger.frommxc(COD1))

#with decomp
(with.decomp_edag <- sum(results_edag))
#error
with.decomp_edag - original_edag
```

Do some data handling to have a nice dataset and plotting

```{r}
results_edag <- data.frame(results_edag)
colnames(results_edag) <- cause_names
results_edag$age <- 16:109
results_edag <- gather(data = results_edag,key = cause,value = contribution,-age)

#now graph results

ggplot(data=results_edag, aes(x=as.factor(age), y=contribution, fill=cause)) +
  ggtitle(bquote(~'Change in '~ e[15]^"\u2020" ~'2005-2015' )) +
  geom_bar(stat = "identity", position = "stack")
```

Let's do the same thing with the stepwise replacement method

First, life expectancy
```{r}
results_step <- stepwise_replacement(func = e0.frommxc, pars1 = c(COD1), pars2 = c(COD2))

#Go back to a matrix
dim(results_step) <- dim(COD1)

#original
(original <- e0.frommxc(COD2) - e0.frommxc(COD1))

#with decomp
(with.decomp <- sum(results_step))

#error
with.decomp - original

results_step <- data.frame(results_step)
colnames(results_step) <- cause_names
results_step$age <- 16:109
results_step <- gather(data = results_step,key = cause,value = contribution,-age)

#now graph results
ggplot(data=results_step, aes(x=as.factor(age), y=contribution, fill=cause)) +
  ggtitle(bquote(~'Change in '~ e[15] ~'2005-2015' )) +
  geom_bar(stat = "identity", position = "stack")

```


Now for lifespan disparity

```{r}
#Now we can perfom the decomposition
results_step_edag <- stepwise_replacement(func = edagger.frommxc, pars1 = c(COD1), pars2 = c(COD2))

#Go back to a matrix
dim(results_step_edag) <- dim(COD1)

#original
(original_edag <- edagger.frommxc(COD2) - edagger.frommxc(COD1))

#with decomp
(with.decomp_edag <- sum(results_step_edag))
#error
with.decomp_edag - original_edag

results_step_edag <- data.frame(results_step_edag)
colnames(results_step_edag) <- cause_names
results_step_edag$age <- 16:109
results_step_edag <- gather(data = results_step_edag,key = cause,value = contribution,-age)

#now graph results
ggplot(data=results_step_edag, aes(x=as.factor(age), y=contribution, fill=cause)) +
  ggtitle(bquote(~'Change in '~ e[15]^"\u2020" ~'2005-2015' )) +
  geom_bar(stat = "identity", position = "stack")
```
